---
title: "Thesis: Exploratory data analysis of use of the Digital Care Guide"
author: "Klara Krøyer Fomsgaard"
date: "2025-09-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Install packages
```{r}
pacman::p_load(tidyverse, 
               dplyr,
               lme4,
               viridis,
               hrbrthemes,
               forcats, 
               MuMIn,
               ggeasy,
               showtext,
               paletteer,
               DHARMa,
               performance,
               emmeans,
               ggeffects,
               glmmTMB,
               colorspace,
               arm)
```

## INITIAL DATA PREP
This section contains the initial data cleaning and preprocessing for explorative analysis

### Load user data
```{r}
data <- read_csv2("app_data.csv")
```
### Data cleaning
```{r}
data_clean <- data %>%
  mutate(times_seen = ifelse(subtask_completed == "No",0,times_seen), #Fill out NAs
         on_time = ifelse(subtask_completed == "No", "Not opened", on_time), #Fill out NAs
         subtask_completed = ifelse(Task_viewed == "No","No", subtask_completed),
         age = case_when(age == 1 ~ "Age 10-19", #Change values for clarity
                                  age == 2 ~ "Age 20-29",
                                  age == 3 ~ "Age 30-39",
                                  age == 4 ~ "Age 40-49",
                                  age == 5 ~ "Age 50-59",
                                  age == 6 ~ "Age 60-69",
                                  age == 7 ~ "Age 70-79",
                                  age == 8 ~ "Age 80-89",
                                  age == 9 ~ "Age 90-99",))

# --- Subset of data excluding NAs
data_subset <- data_clean %>% 
  filter(!is.na(Task_viewed)) %>%
  filter(!is.na(subtask_completed))

```

## DEMOGRAPHICS
This section provides descriptive statistics of the demogagraphic distribution of app users, as well as a plot showing the distribution
```{r}
# --- Make list of users with attached demographics
demographics <- data_subset %>% 
  group_by(ID,age,sex) %>% 
  summarise() %>% 
  filter(is.na(ID) == FALSE)  # Remove unspecified participant
```
```{r}
# --- Calculate number of women and men
demographics %>% 
  group_by(sex) %>% 
  count()
```

```{r}
# --- Calculate how many percent of the users are within 50-79
demographics_grouped <- demographics %>% 
  group_by(age) %>% 
  summarise(n = n()) 

(sum(c(29,41,38))/sum(demographics_grouped$n))*100
```

### Visualization of demographics
```{r}
# --- Import design features
font_add(family = "Sofia Pro",
         regular = "/Users/klarafomsgaard/Downloads/Sofia/Sofia Pro Regular Az.otf")
font_add(family = "Sofia Pro Bold",
         regular = "/Users/klarafomsgaard/Downloads/Sofia/Sofia Pro Bold Az.otf")
showtext_auto()

# --- Color palettes
pal <- paletteer_d("fishualize::Lampris_guttatus")
## Extend palette 
pal_extended <- c(pal, "#999999", "#000000")
  
```

```{r}
# --- PLOT: Visualize age and sex distribution
ggplot(demographics, aes(x = age)) +
  geom_bar(aes(fill = sex)) +
  ggtitle("Demographical data for app users", subtitle = "Distribution across age intervals and sex") +
  # Visual aesthetics
  theme_minimal() +
  scale_fill_paletteer_d("fishualize::Lampris_guttatus") +
  theme(
    text = element_text(family = "Sofia Pro"),
    plot.title = element_text(family = "Sofia Pro Bold", hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )+
  labs(x = "Age intervals", y = "Number of users")+
  coord_cartesian(ylim = c(0, 45)) 
```
```{r}
# --- PLOT: Visualize distribution of male and female users
ggplot(demographics, aes(x = sex)) +
  geom_bar(aes(fill = sex)) +
  # Visual aesthetics
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  ggtitle("Demographics", subtitle = "Distribution across sex") +
  theme_minimal() +
  scale_fill_paletteer_d("fishualize::Lampris_guttatus")+
  theme(
    text = element_text(family = "Sofia Pro"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5))+
  labs(x = "Sex", y = "Number of users")
```

## INVESTIGATING TASK ENGAGEMENT
This section investigates task engagement through data visualization and descriptive statistic. The section also includes a post hoc analysis of trend in engagement across ages. 

### Overview of subtask completion distribution
```{r}
# --- Make binary category for whether a task was completed or not
data_bin <- data_subset %>% 
  mutate(subtask_completed = case_when(
    subtask_completed == "Yes" ~ 1,
    subtask_completed == "No" ~ 0),
    times_seen = ifelse(subtask_completed == 0, 0,times_seen
  ))

## Convert to factor
data_bin$age <- as.factor(data_bin$age)

## Summarize number of completed tasks and engagement rate for each user
subtask_dist <- data_bin %>%
  group_by(ID, subtask_completed) %>%
  count(name = "n") %>%   # Count data                        
  ungroup() %>%                                   
  complete(ID, subtask_completed = c(0, 1), fill = list(n = 0)) %>% # Calculate engagement rate
  mutate(percentage = (n / 27) * 100) %>%
  filter(subtask_completed == 1 & is.na(ID) == FALSE)

# --- Calculate mean and SD
## Count data
mean(subtask_dist$n)
sd(subtask_dist$n)

## Percentage
mean(subtask_dist$percentage)
sd(subtask_dist$percentage)

# --- HISTOGRAM for task engagement
ggplot(subtask_dist) +
  geom_histogram(aes(x = percentage,fill = 'green'), 
           binwidth = 1)  + 
  #  Visual aesthetics
  theme_minimal()+
  ggtitle("Distribution of subtask engagement among users", subtitle = "% of subtasks opened") +
  labs(x = "Subtask engagement in %", 
       y = "Number of users",
       fill = NULL) + 
  scale_fill_paletteer_d("fishualize::Lampris_guttatus")+
  theme(legend.position = 'null',
        text = element_text(family = "Sofia Pro"),
        plot.title = element_text(family = "Sofia Pro Bold", hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

### Visualizing engagement rate for each age group
```{r}
# --- Append age
completion_grouped_age <- subtask_dist %>% 
  merge(demographics, by = "ID") %>% 
  filter(age != "Age 10-19")

# --- BOXPLOT: Task engagement across age groups
ggplot(completion_grouped_age, aes(x = factor(age), y = percentage, fill = factor(age))) +
  geom_jitter(width = 0.1, alpha = 0.7, size = 2, aes(color = age)) +
  geom_boxplot(alpha = 0.6, aes(color = after_scale(colorspace::darken(fill, 0.5)))) +
  # Visual aesthetics
  labs(x = "Age", y = "% completed") +
  theme_minimal() +
  scale_fill_manual(values =pal_extended)+
  scale_color_manual(values =pal_extended)+
  ggtitle("Subtask engagement across age groups", subtitle = "Completion rate out of total number of subtasks in %") +
  theme(legend.position = 'null',
        text = element_text(family = "Sofia Pro"),
        plot.title = element_text(family = "Sofia Pro Bold", hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, margin = margin(b = 10),size = 10),
        axis.title.x = element_text(margin = margin(t = 8)),
        axis.title.y = element_text(margin = margin(r = 8)))

```

#### Post Hoc Analysis: Trend in engagement across ages
Due to an indication of a negative trend across age in the visual inspection a post hoc analysis is conducted to test for significance. Bootstrapping is inlcuded to check for robustness, despite difference in sample size.
```{r}
# --- Treat age groups as ordinal variable
data_bin <- data_bin %>%
   mutate(age_ordinal = as.numeric(factor(age, ordered = TRUE)))


# --- MIXED EFFECTS LOGISTIC REGRESSION
model <- glmer(subtask_completed ~ age_ordinal + (1|ID),
               data = data_bin, family = binomial)

## Review results
summary(model)
```

```{r}
# --- BOOTSTRAPPING: Testing for robustness of significance
set.seed(123)

n_boot <- 100  # number of bootstrap iterations (use smaller number first)
participants <- unique(data_bin$ID)
boot_coefs <- numeric(n_boot)

for (i in 1:n_boot) {
  # Resample participants with replacement
  sampled_IDs <- sample(participants, length(participants), replace = TRUE)
  
  # Subset data to include only sampled participants
  boot_data <- do.call(rbind, lapply(sampled_IDs, function(id) data_bin[data_bin$ID == id, ]))
  
  # Fit the same glmer model
  model_boot <- glmer(subtask_completed ~ age_ordinal + (1|ID),
                      data = boot_data, family = binomial)
  
  # Save coefficient of interest (age_ordinal)
  boot_coefs[i] <- fixef(model_boot)["age_ordinal"]
}

## Get bootstrap 95% confidence interval
quantile(boot_coefs, c(0.025, 0.975))
```


### Visualizing completion across different subtasks
```{r}
# --- Import subtask key
subtask_key <- read_csv2("subtask_key.csv")
```

```{r}
# --- DATA PREP
# --- Summarise how many users completed each subtask
subtasks_acc <- data_bin %>% 
  group_by(subtask) %>% 
  count(subtask_completed)

subtasks_acc$subtask_completed <- fct_rev(as.factor(subtasks_acc$subtask_completed)) 

# --- Calculate percentage
subtasks_acc_compl <- subtasks_acc %>% 
  group_by(subtask) %>% 
  mutate(total_completed = sum(n)) %>% 
  ungroup() %>% 
  mutate(percentage_completed = (n/total_completed)*100) %>%
  filter(subtask_completed == 1)

# ---  Merge with key
subtasks_acc_compl <- merge(subtasks_acc_compl,subtask_key, by = "subtask")

# --- Extract overall task
subtasks_acc_compl <- subtasks_acc_compl %>%
  mutate(
    task_number = str_extract(subtask_description, "(?<=Task\\s)\\d+"),
    task_description = str_remove(subtask_description, "Task\\s*\\d+\\s*\\-")
  )

# --- Calculate mean and SD
mean(subtasks_acc_compl$percentage_completed)
sd(subtasks_acc_compl$percentage_completed)
# --- Calculate min & max
min(subtasks_acc_compl$percentage_completed)
max(subtasks_acc_compl$percentage_completed)


```

```{r}
# --- DATA PREP
df <- subtasks_acc_compl %>% 
  mutate( sub_unique = paste0(task_description, "___", row_number()), 
          sub_unique = factor(sub_unique, levels = rev(sub_unique))) 

#--- PLOT: Engagement with subtasks
ggplot(df, aes(x = percentage_completed, 
               y = sub_unique, 
               fill = percentage_completed)) + 
  # Visual aesthestics
  geom_col(color = "darkgrey", size = 0.1, width = 0.8, alpha = 0.9) +
  xlim(0,100)+
  scale_x_continuous(
    limits = c(-1.5, NA),          
    expand = expansion(mult = 0)
  )+
  scale_y_discrete(labels = ~ gsub("___\\d+$", "", .x)) +
  scale_fill_distiller(palette = "Spectral", direction = 1) +
  labs(x = "% of users", y = NULL, fill = "%") + 
  theme_minimal() + 
  ggtitle("Engagement with subtasks",subtitle = "% of total number of users who have opened the subtask")+
  theme(
        legend.position = "none",
        text = element_text(family = "Sofia Pro"),
        plot.title = element_text(family = "Sofia Pro Bold", hjust = 0.5, face = "bold"), 
        plot.subtitle = element_text(size = 9, hjust = 0.5, margin = margin(b=10)),
        plot.title.position = "plot", 
        axis.text.y = element_text(size = 8, margin = margin(r = 1.5)),
        axis.ticks.length = unit(1, "pt"),
        axis.title.x = element_text(margin = margin(t = 10)),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 10),
        legend.key.size = unit(0.5, "cm"),
        )
   
```

### Visualizing whether users complete the task on time
The data is explored looking only and users engaging with the tasks
```{r}

# --- DATA PREP
on_time_acc <- data_bin %>% 
  filter(!is.na(on_time)) %>% 
  filter(subtask_completed == 1) %>% 
  group_by(subtask) %>% 
  count(on_time) %>% 
  merge(subtask_key, by = "subtask") %>% 
  mutate(
    task_number = str_extract(subtask_description, "(?<=Task\\s)\\d+"),
    full_description = subtask_description,
    subtask_description = str_remove(subtask_description, "Task\\s*\\d+\\s*\\-"),
    subtask = factor(subtask)) %>% 
  group_by(subtask) %>% 
  mutate(percentage = (n/sum(n))*100)


# --- PLOT: Timing of engagement
ggplot(on_time_acc, aes(x = n, 
               y = fct_reorder(full_description, as.numeric(subtask),.desc = TRUE), 
               fill = on_time)) + 
  geom_col(size = 0.2, 
           width = 0.8, 
           position = position_dodge()) +
  # Visual aesthetics
  xlim(0,100)+
  scale_x_continuous(
    limits = c(-1.5, NA),          
    expand = expansion(mult = 0)) +
  scale_y_discrete(labels = ~ gsub("Task \\d+ - ", "", .x)) +
  scale_fill_manual(values = c("#F75107FF","#85D3C3FF")) +
  labs(x = "% of users", y = NULL, fill = "on time") + 
  theme_minimal() + 
  ggtitle("Timing of subtask engagement",subtitle = "Including only users who have opened the subtask")+
  theme(
        text = element_text(family = "Sofia Pro"),
        plot.title = element_text(family = "Sofia Pro Bold", hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(size = 9, hjust = 0.5, margin = margin(b=10)), 
        plot.title.position = "plot", 
        axis.text.y = element_text(size = 8, margin = margin(r = 1.5)),
        axis.ticks.length = unit(1, "pt"),
        axis.title.x = element_text(margin = margin(t = 10)),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.5, "cm"))

```

### Visualization of whether tasks have been marked as complete
This section investigate behavior associated with the 'Mark as completed' feature, which users can use to track progress
```{r}
# --- Import task key
task_key <- read_csv2("task_key.csv")

# --- Extract the number of users who have opened the tasks to overlay on plot
task_opened_subset <- data_task_acc %>% 
  filter(Task_viewed == "Yes") %>% 
  merge(task_key,by = "task")

# --- DATA PREP
data_marked <- data_bin %>%
  mutate(marked_as_completed = ifelse(is.na(marked_as_completed),"No", marked_as_completed),
         marked_as_completed = ifelse(marked_as_completed == "yes", "Yes", marked_as_completed),
         marked_as_completed = ifelse(Task_viewed == "No", "Not opened", marked_as_completed)) %>%
  filter(Task_viewed == "Yes") %>% 
  group_by(ID, task) %>%      # group by participant ID and task
  slice_head(n = 1) %>%       # keep only the first row in each group
  ungroup() %>% 
  group_by(task) %>%
  count(marked_as_completed) %>% 
  merge(task_key, by = "task") 

data_marked_yes <- data_marked %>% 
  filter(marked_as_completed == "Yes")

# --- PLOT: Combining the total number of active users for each task and the amount of users opening the task
ggplot(data_marked_yes, 
       aes(x = n, 
           y = fct_reorder(task_description, as.numeric(task),.desc = TRUE),
           fill = marked_as_completed)) + 
  # Total number of active users
  geom_col(data = task_opened_subset, 
           aes(x = n, 
               y = fct_reorder(task_description, as.numeric(task), .desc = TRUE)),
           fill = "black", 
           width = 0.05, inherit.aes = FALSE) +
  geom_point(aes(x = task_opened_subset$n, 
               y = fct_reorder(task_opened_subset$task_description, 
                               as.numeric(task_opened_subset$task), 
                               .desc = TRUE),
               color = "Total interactions"),
             shape = 108, size = 5) +
  # Number of users marking the task as completed
  geom_col(size = 0.2, 
           width = 0.7) +
  # Visual aesthetics
  theme_minimal() +
  ggtitle("Feature engagement: Marking tasks as completed") +
  labs(x = "Number of users", 
       y = NULL, 
       fill = "Users marking task as completed",
       color = "Total user interactions") + 
    scale_x_continuous(
    limits = c(-1.5, 120),
    expand = expansion(mult = 0)
  )+
  scale_fill_manual(values = pal[2])+
  scale_color_manual(values ="black")+
  theme(
        text = element_text(family = "Sofia Pro"),
        plot.title = element_text(family = "Sofia Pro Bold", hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 9, margin = margin(b=10)),
        plot.title.position = "plot",
        axis.text.y = element_text(size = 9, margin = margin(r = 1.5)),
        axis.ticks.length = unit(1, "pt"),
        axis.title.x = element_text(margin = margin(t = 10)),
        legend.text = element_text(size = 0),
        legend.title = element_text(size = 8),
        legend.key.size = unit(0, "cm"),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.justification = "center",              
        legend.box.just = "center")

```

